A state space model reveals the rich and varied dynamics of hippocampal replay

### 2020-07-18 Article.tex
Update on Overleaf.

\title{A state space model reveals the rich and varied dynamics of replay}
\shorttitle{Replay Dynamics}

@@

Mental events may not occur in the same time span as actual events; memories can be retrieved and used in less time than was required for the original experience, and mental simulations can span a series of actions more quickly than would be required to perform those actions. Consistent with this, generative representations can unfold rapidly. As a result, decoding generative representations often involves making assumptions about their timescale and temporal structure, and these assumptions can limit the interpretation of the data.

@@

Identifying these events typically involves multiple steps and assumptions about the nature of the event. First, an encoding model is constructed based on spiking activity during movement, most often using only spikes that have been clustered as single units (putative single neurons). Then, a subset of SWRs or events with high activity levels are selected based on an threshold for event size chosen by the experimenter. The spikes within these events decode the probability of position in overlapping or non-overlapping time bins whose size is also chosen by the experimenter. Finally, the most commonly used approach to detecting sequential replay involves fitting a line to the resulting set of probability distributions, which instantiates the assumption that the representations progress at a constant speed. A statistical test is then used to determining whether the fit is better than the fit to shuffled versions of the data.

While this approach identifies constant speed events, it does not consider events that are rejected by the statistical test, and the use of a fixed size temporal bin acts as a boxcar smoother that limits the potential movement speeds of the representation. The linear fit is also problematic, because it has the potential to reject real events that do not move at constant speeds. Moreover, there is now evidence that a subset of these events have stationary representations of a single location, which are often excluded. In addition, these methods identify only a subset of events as containing content (typically somewhere between 10 and 45\% of the chosen events), and thereby provide little insight into the remaining majority of the events.

 Recognizing the problems with the linear fit, several studies have moved away from the constant velocity assumption, using the weighted mean or maximum of each posterior time bin and connecting each time bin with a line. For example, using this approach, Pfeiffer and Foster \cite{PfeifferAutoassociativedynamicsgeneration2015} found that replays can alternate between representing a single location and sequential spatial trajectories. On the other hand, Stella et al. \cite{StellaHippocampalReactivationRandom2019} found that replays are more spatially continuous, following Brownian diffusion dynamics. Both methods still used large time bins and neither took into account the uncertainty of the decoded estimates, making it hard to identify the source of the different conclusions.

@@

We therefore developed a state space model of generative activity that achieves all of those goals. State space models are a well-understood, well-known statistical solution to the problems described above. By mathematically modeling the relationship between the data and latent dynamics, state space models make the assumptions of the model explicit and interpretable. Our model goes beyond previous approaches \cite{MaboudiUncoveringtemporalstructure2018, DengRapidclassificationhippocampal2016} and characterizes represented trajectories as a mixture of three underlying patterns of dynamics: stationary trajectories, continuous trajectories that are many times the typical speed of the animal, and spatially fragmented trajectories. We show how this model can take advantage of clusterless decoding---which relates multiunit spike wave form features to position without spike sorting---giving us more information about the population spiking activity. We apply this model to data from 10 animals, focusing on SWR events to permit a direct comparison to previous work. We find that the vast majority of SWRs contain spatially coherent content. Surprisingly, while the expected high speed replay events were identified, the most common category of events expressed representations that moved at speeds consistent with real animal movement, suggesting that high speed replay is not the most common manifestation of generative activity. Our method provides a useful and powerful tool for understanding the dynamics of mental representations, and could be adapted for use in many brain areas.

@@

Finally, we specify how likely each dynamic is to persist in time versus changing to another dynamic (Figure \ref{1}D). In order to be conservative, we assume that each dynamic is likely to dominate for the duration of the SWR, with a small probability of switching to one of the other dynamics (note that below we show that our results are relatively insensitive to this parameter). Accordingly, we set the probability of staying in a state to 0.98 for each 2 ms time step. Because the probability of changing dynamics follows a geometric distribution, this corresponds to an expected duration of 100 ms for staying in a particular state.

@@

We note here that the choice of any particular threshold is arbitrary, and that the power of our approach lies in part in the ability to assign a confidence value for each dynamic or combinations of dynamics to each moment in time. Our goal in choosing 0.8 was to use a threshold that corresponds to highly probable and to explicitly avoid the arbitrary choice of 0.95 because the 0.95 threshold often leads people to conclude that only events that are statistically significant at the 0.05 level are likely to be meaningful to the brain. Nonetheless, we also verified that our central results hold with a higher threshold of 0.95.

As mentioned above, the model is robust to the choice of probability of staying in a state. We decoded the spiking sequence in Figure \ref{1}A with different probabilities of staying in the same state versus switching to another state (Figure 1-Supplemental Figure 1). We found that for a large range of plausible probabilities of staying in one of the dynamics (0.9 - 0.9999), the model still correctly identified the dynamic with high probability.

@@

Previous work focusing on spatially sequential replay reported that only a minority of events contain sequential spatial content [REFS]. We therefore asked what fraction of classified events contained spatially coherent content in any form, defined as a SWR containing any times with stationary, stationary-continuous mixture, or continuous dynamics (see Methods). We found that 22478 of 23382 or 96\% of classified SWRs had time points with spatially coherent structure. We then asked what fraction of events contained spatially incoherent content, defined as a SWR containing any times with fragmented-continuous mixture or fragmented. We found that only 3353 of 23382 or 14\% of classified SWRs had time points with spatially incoherent structure. This was consistent across animals (Figure \ref{3}F). We then asked what percentage of classified SWR contained continuous content, like would typically be analyzed when using the Bayesian decoder. Here our findings were consistent with previous reports: in our datasets 4511 of 23382 or 19\% of classified SWRs had some continuous content (Figure \ref{3}G). Thus, focusing on only continuous content misses a large fraction of events where there is evidence for spatially coherent, but not continuous content.

We repeated this analysis with a higher threshold of 0.95 to ensure that our result was not dependent on this threshold. We found that, while this slightly reduced the total fraction of classified ripples (19180 / 25845 or 74\% SWRs), an even higher fraction of the classified ripples (19020 / 19180 or 99\% classified) had spatially continuous content. Classified SWRs containing spatially incoherent content similarly were a small fraction of the classified SWRs (spatially incoherent: 484 / 19180 or 3\%).

@@

We found that spatially coherent events did indeed have lower cumulative HPDs. Figure \ref{4}A and B show two example SWRs that were classified as having stationary and continuous dynamics, respectively. The bulk of the HPD values at each time step in these SWRs is concentrated in a small portion of the track and the cumulative HPD is low throughout the SWRs. In contrast, Figure \ref{4}C shows a SWR where the dynamics are fragmented and correspondingly, the HPD is much more spatially diffuse and the cumulative HPD is much higher. The HPD also provides insights into the moment-by-moment structure of each event, which can change over the time course of a SWR. An example of this change is shown in Figure \ref{4}D, where the cumulative HPD is small for most of the SWR until the end, where the uncertainty of position becomes much higher, reflecting a transition from a spatially coherent to a spatially incoherent representation. Overall, when we examined the average cumulative HPD for each SWR, grouped by dynamic, we found a clear bi-modal split between spatially coherent dynamics and spatially incoherent dynamics (Figure \ref{4}E). For the spatially coherent dynamics, the cumulative HPD was much lower than the spatially incoherent dynamics (median 24 cm, spatially coherent vs. median 238 cm, spatially incoherent).

@@

Our results indicate that most events with coherent spatial content are not continuous; that is, trajectories that move many times the running speed of the animal. We therefore examined these events in more detail. We first noted that most of the SWRs (16454 of 23382 classified SWRs or 70\%) were well described by a single dynamic. Surprisingly, the most common SWRs were categorized exclusively as stationary-continuous mixtures (9860 of 23382 classified SWRs or 42\%, Figure \ref{5}A). These events contain representations that move at rates more consistent with real-world movement (Fig. 1G), but were slightly shorter, on average, than events with continuous dynamics (median duration: stationary-continuous-mixture 72 ms vs. continuous 95 ms, Figure \ref{5}B). Nonetheless, both these slow-moving events and continuous events represented locations that were some distance away from the animal (mean trajectory distance from animal's position: stationary-continuous-mixture 51 cm vs. continuous 42 cm, Figure \ref{5}C). This indicates that the trajectories of these SWRs, like ones that have faster continuous dynamics, do not represent the animal's actual position and cannot be explained by erroneously detected SWRs.

@@

\section*{Discussion}
Here we developed a state space model that identifies and quantifies generative activity in terms of a mixture of three movement dynamics: stationary, continuous, and fragmented. We showed that this model is interpretable; it has well-defined parameters and allows us to explain a large fraction of SWRs in terms of these dynamics---far greater than had been explained before. We showed that this model is flexible; it can be used with sorted or clusterless spikes, it can be used with 1D or 2D positions, and---because it uses a mixture of dynamics---it allowed us to confidently discover both SWRs with continuous many times the speed of the animal and SWRs with non-constant speed and non-linear trajectories that would not be found with the typical Bayesian decoder. Finally, we showed that this model is robust; it was relatively insensitive to misspecified transition parameters between dynamics and depended strongly on the hippocampal place representation structure.

Using this model, we found that most SWRs contained spatially coherent trajectories. Importantly, this content was not predominantly trajectories that occurred at many times the speed of the animal, but trajectories that were much closer to the animal's actual movement speed. Many of these events did not represent the animal's current location, but other locations on track. These results challenge the long-held notion that hippocampal replay trajectories necessarily proceed at many times the speed of the animal \cite{NadasdyReplayTimeCompression1999, LeeMemorySequentialExperience2002, DavidsonHippocampalReplayExtended2009}. Instead there is a much richer variety of speeds the animal can use to recall prior experience or generate possible future events, including stationary representations of single locations.

Our results also explain the seeming conflict between Stella and colleagues \cite{StellaHippocampalReactivationRandom2019}, who found that hippocampal replay trajectories followed Brownian-like diffusion dynamics, and Pfeiffer and Foster \cite{PfeifferAutoassociativedynamicsgeneration2015}, who found that replay trajectories contained jumps in position. We found that while replay speeds are on average distributed log normally, there are still definitive "jumps" from one position to another in a small subset of the replays. Our model allows identification of both types of sequences.

We were able to find these results for several key reasons. First, we sought to describe as many SWRs as possible, instead of throwing many events out. Second, we used clusterless decoding, which enabled us to take advantage of information from neurons farther away from the tetrodes and not just those that could be cleanly and somewhat subjectively separated. Third, we built an explicit, but flexible model that allowed us to interpret more of the SWRs. And fourth, we used small time bins and acausal estimation of latent position, which allows us to identify the specific timing of the dynamics.

Our model can be viewed as an extension to the multiple temporal models approach of Johnson and colleagues 2008 \cite{JohnsonMeasuringdistributedproperties2008}, which used model selection to determine the model with the most appropriate trajectory speed. Our model goes beyond the Johnson model in that it explicitly permits a mixture between the movement speeds, can work for arbitrary track geometries, and uses clusterless decoding. Our model also represents a middle ground between Hidden Markov-style models \cite{MaboudiUncoveringtemporalstructure2018, ChenBayesiannonparametricmethods2016, LindermanBayesiannonparametricapproach2016, ChenUncoveringspatialtopology2012}, which seek to be environment-agnostic discovers of sequential patterns and the typical Bayesian decoder, which either uses very restrictive assumptions about the nature of the trajectories or no assumptions. The Hidden Markov-style models require lots of data and still involve interpreting the sequences in terms of known variables, such as the environment. The Bayesian decoder is either quite restrictive, when assuming a constant velocity trajectory, or potentially quite noisy, depending on the data. Both of these styles of models are dependent on the time bin size. Our model leverages some of our prior knowledge about hippocampus and its relationship to space to give less noisy and more interpretable estimates of the replay trajectories.

We believe our model can enable a richer set of analyses that takes advantage of all the data. The model can easily be extend to incorporate other previously experienced environments by training place field models on those environments and including the appropriate movement transition matrices for those environments.

@@

\subsection*{Simulated Data}
Encoding data for Figure \ref{1} and Figure \ref{fig:Figure1-Figure supplement 1} was generated by simulating 15 traversals of a 180 cm linear track. The spiking of 19 neurons was simulated using an inhomogeneous Poisson process where the instantaneous firing rate of each neuron changes according to a Gaussian place field. The Gaussian place fields had a variance of 36 cm and were spaced at 10 cm intervals over the 180 cm track. The decoding data for Figure \ref{1}G was generated by simulating 20,000 linear traversals of the 180 cm track, each with a unique constant speed, starting at 0.5 cm/s and increasing by 0.5 cm/s up to 10,000 cm/s. Each simulated neuron "spiked" when the traversal passed through the peak location of its place field.

@@

\subsection*{Behavioral Task}
All animals performed a W-track spatial alternation task, which is described in detail in \cite{KarlssonAwakereplayremote2009}. In brief, each day, animals alternated between 20 minute rest epochs in a rest box and 15 minutes run epochs in the W-track. On the W-track, animals alternated between two types of runs: inbound and outbound. On inbound runs, the animal had to run from an outer arm to the center arm. On outbound runs, the animals had to run from the center arm to the least recently visited outer arm. Animals were rewarded for performing the task correctly. Only run epochs with at least 9 putative hippocampal pyramidal cells that fired at least 100 spikes were included in the analysis.

@@

\begin{equation*}
  p(x_{k} \mid x_{k-1}, I_{k}, I_{k-1}) = 
  \begin{blockarray}{*{3}{c} l}
    \begin{block}{*{3}{>{$\footnotesize}c<{$}} l}
      $I_{k} = Stationary$ & $I_{k} = Continuous$ & $I_{k} = Fragmented$ \\
    \end{block}
    \begin{block}{[*{3}{c}]>{$\footnotesize}l<{$}}
      \delta(x_{k-1}) & \mathcal{N}(x_{k-1}, 6.0) & \mathcal{U}(\min x, \max x) \bigstrut[t]& $I_{k-1} = Stationary$ \\
      \delta(x_{k-1}) & \mathcal{N}(x_{k-1}, 6.0)  & \mathcal{U}(\min x, \max x) & $I_{k-1} = Continuous$ \\
      \mathcal{U}(\min x, \max x) & \mathcal{U}(\min x, \max x) & \mathcal{U}(\min x, \max x) & $I_{k-1} = Fragmented$ \\
    \end{block}
  \end{blockarray}
\end{equation*}
where $\delta(x_{k-1})$ is an identity transition matrix where position cannot change from the previous time step, $\mathcal{N}(x_{k-1}, 6.0)$ is a random walk from the previous position with variance 6.0, and $\mathcal{U}(\min x, \max x)$ is a uniform transition that allows transitions to any possible position. As discussed in the Results, this says that when persisting in the same dynamic, the stationary, continuous, and fragmented dynamics are defined by the identity transition, the random walk, and the uniform transition, respectively. When transitioning to or from the fragmented dynamic, we assume we do not have any information about the position, so the transition is uniform. Finally, when the transition is from the stationary to continuous, we assume the position is spatially close where it was previously, so we use a random walk. When the transition is from continuous to stationary, we assume that the position is no longer changing, so we use the identity transition.

@@

\subsection*{Encoding - Clusterless}
In order to encode how each tetrode's unsorted spiking activity and waveform features relate to position, we used kernel density estimation (KDE) to estimate the following distributions: $\pi(x)$, $p_{i}(x_k)$, and $p_{i}(x_k, \vec{m}_k)$. We used KDEs of the form: 
$$
kde(y) = \frac{1}{N h_1...h_D} \sum^{N}_{i=1} \prod^{D}_{d=1} K_{d}\left(\frac{y_d - y_{i,d}}{h_d}\right)
$$
where $y$ is the data with $D$ dimensions, $K$ is a 1 dimensional Gaussian kernel with bandwidth $h_d$, $y_{i, d}$ is the data observed during movement, $N$ is the number of observations during the movement period. For $\pi(x)$, $y_{i, d}$ is all positions observed during movement. For $p_{i}(x_k)$, $y_{i, d}$ is all positions at the time of multiunit spikes during movement. For $p_{i}(x_k, \vec{m}_k)$, $y_{i, d}$ is all positions at the time of multiunit spikes and their associated waveform features during movement. We choose the bandwidth $h_d$ to be 6.0 cm for any position dimension and 24.0 $\mu V$ for any spike amplitude dimension, because these parameters minimized decoding error in \cite{KloostermanBayesiandecodingusing2014}. The mean firing rate $\mu_k$ was also estimated during movement.

>>>>

\title{A state space model reveals the rich and varied dynamics of hippocampal replay}
\shorttitle{Replay Dynamics}

@@

Mental events may not occur in the same time span as actual events: memories can be retrieved and used in less time than was required for the original experience, and mental simulations can span a series of actions more quickly than would be required to perform those actions. Consistent with this, generative representations can unfold rapidly. As a result, decoding generative representations often involves making assumptions about their timescale and temporal structure, and these assumptions can limit the interpretation of the data.

@@

Identifying these events typically involves multiple steps and assumptions about the nature of the event. First, an encoding model is constructed based on spiking activity during movement, most often using only spikes that have been clustered as single units (putative single neurons). Then, a subset of SWRs or events with high activity levels are selected based on an threshold for event size chosen by the experimenter (cite Foster and Wilson, Diba and Buzsaki, Karlsson and Frank, appropriate Csicsvari paper). The spikes within these events decode the probability of position in overlapping or non-overlapping time bins whose size is also chosen by the experimenter. Finally, the most commonly used approach to detecting sequential replay involves fitting a line to the resulting set of probability distributions, which instantiates the assumption that the representations progress at a constant speed (Foster and Wilson, Diba and Buzsaki, Karlsson and Frank,). A statistical test is then used to determining whether the fit is better than the fit to shuffled versions of the data.

While this approach identifies constant speed events, it does not consider events that are rejected by the statistical test, and the use of a fixed size temporal bin acts as a boxcar smoother that limits the potential movement speeds of the representation. The linear fit is also problematic, because it has the potential to reject real events that do not move at constant speeds. Approaches that focus on the order of cell activity within each even (Lee and Wilson, Gupta et. al?) allow for a relaxation of that linear assumption, but replace by either ignoring all but the first spike from each cell or some arbitrarily smoothed version of the spike train. These approaches also do not provide information about the dynamics of the underlying spatial representation. Moreover, there is now evidence that a subset of these events have stationary representations of a single location, which both approaches expclude . In addition, these methods identify only a subset of events as containing content (typically somewhere between 10 and 45\% of the chosen events), and thereby provide little insight into the remaining majority of the events.

 Recognizing the problems with the linear fit, several studies have moved away from the constant velocity assumption, using the weighted mean or maximum of each posterior time bin and connecting each time bin with a line. For example, using this approach, Pfeiffer and Foster \cite{PfeifferAutoassociativedynamicsgeneration2015} found that replays can alternate between representing a single location and sequential spatial trajectories. On the other hand, Stella et al. \cite{StellaHippocampalReactivationRandom2019} found that replays are more spatially continuous, following Brownian diffusion dynamics. Both methods still used large time bins and neither took into account the uncertainty of the decoded estimates, however, making it hard to identify the source of the different conclusions.

@@

We therefore developed a state space model of generative activity that achieves all of those goals. State space models are a well-understood, well-known statistical solution to the problems described above. By mathematically modeling the relationship between the data and latent dynamics, state space models make the assumptions of the model explicit and interpretable. Our model goes beyond previous approaches \cite{MaboudiUncoveringtemporalstructure2018, DengRapidclassificationhippocampal2016} and characterizes represented trajectories as a mixture of three underlying patterns of dynamics: stationary trajectories, continuous trajectories that are many times the typical speed of the animal, and spatially fragmented trajectories. We show how this model can take advantage of clusterless decoding---which relates multiunit spike wave form features to position without spike sorting---giving us more information about the population spiking activity. We apply this model to data from 10 animals, focusing on SWR events to permit a direct comparison to previous work.

We find that the vast majority of SWRs contain spatially coherent content. Surprisingly, while the expected high speed replay events were identified, the most common category of events expressed representations that moved at speeds consistent with real animal movement, suggesting that high speed replay is not the most common manifestation of generative activity. These findings illustrate the power of state space models and provide a new understanding of the nature of hippocampal replay.

@@

Finally, we specify how likely each dynamic is to persist in time versus changing to another dynamic (Figure \ref{1}D). In order to be conservative, we assume that each dynamic is likely to dominate for ~100 ms, which is approximately the duration of the SWR. There is however,  a small probability of switching to one of the other dynamics (note that below we show that our results are relatively insensitive to this parameter). Accordingly, we set the probability of staying in a dynamic to 0.98 for each 2 ms time step. Because the probability of changing dynamics follows a geometric distribution, this corresponds to an expected duration of 100 ms for staying in a particular dynamic. Importantly, the data drive the dynamic, so even if the probability of a particular dynamic is high, data providing clear evidence for a change in dynamic can, within the time frame of ~10 ms, drive a change to a different dynamic.

@@

We note here that the choice of any particular threshold is arbitrary, and that the power of our approach lies in part in the ability to assign a confidence value for each dynamic or combinations of dynamics to each moment in time. Our goal in choosing 0.8 was to use a threshold that corresponds to highly probable and to explicitly avoid the arbitrary choice of 0.95 because the 0.95 threshold often leads people to conclude that only events that are statistically significant at the 0.05 level are meaningful in the brain. Nonetheless, we also verified that our central results hold with a higher threshold of 0.95.

As mentioned above, the model is robust to the choice of probability of staying in a state. To demonstrate this we decoded the spiking sequence in Figure \ref{1}A with different probabilities of staying in the same state versus switching to another state (Figure 1-Supplemental Figure 1). We found that for a large range of plausible probabilities of staying in one of the dynamics (0.9 - 0.9999), the model still correctly identified the dynamic with high probability.

@@

Previous work focusing on spatially sequential replay reported that only a minority of events contain sequential spatial content (Karlsson and Frank, Foster and Wilson Davidson et. al. etc. . We therefore asked what fraction of classified events contained spatially coherent content in any form, defined as a SWR containing any times with stationary, stationary-continuous mixture, or continuous dynamics (see Methods). We found that 22478 of 23382 or 96\% of classified SWRs (85\% of all SWRs) included spatially coherent structure . We then asked what fraction of events contained spatially incoherent content, defined as a SWR containing any times with fragmented-continuous mixture or fragmented dynamics. We found that only 3353 of 23382 or 14\% of classified SWRs had time points with spatially incoherent structure. This was consistent across animals (Figure \ref{3}F).

To better compare our findings to previous work, we quantified the percentage of classified SWRs that contained continuous content, as would typically be analyzed when using the Bayesian decoder. Here our findings were consistent with previous reports: in our datasets 4511 of 23382 or 19\% of classified SWRs had some continuous content (Figure \ref{3}G). Thus, focusing on only continuous content misses a large fraction of events where there is evidence for spatially coherent, but not continuous content.

We repeated this analysis with a higher threshold of 0.95 to ensure that our result was not dependent on this threshold. We found that, while this slightly reduced the total fraction of classified ripples (19180 / 25845 or 74\% SWRs), an even higher fraction of the classified ripples (19020 / 19180 or 99\% classified) had spatially coherent content. Classified SWRs containing spatially incoherent content similarly were a small fraction of the classified SWRs (spatially incoherent: 484 / 19180 or 3\%).

@@

We found that spatially coherent events did indeed have lower cumulative HPDs than events with fragmented dynamics. Figure \ref{4}A and B show two example SWRs that were classified as having stationary and continuous dynamics, respectively. The bulk of the HPD values at each time step in these SWRs is concentrated in a small portion of the track and the cumulative HPD is low throughout the SWRs. In contrast, Figure \ref{4}C shows a SWR where the dynamics are fragmented and correspondingly, the HPD is much more spatially diffuse and the cumulative HPD is much higher. The HPD also provides insights into the moment-by-moment structure of each event, which can change over the time course of a SWR. An example of this change is shown in Figure \ref{4}D, where the cumulative HPD is small for most of the SWR until the end, where the uncertainty of position becomes much higher, reflecting a transition from a spatially coherent to a spatially incoherent representation. Overall, when we examined the average cumulative HPD for each SWR, grouped by dynamic, we found a clear bi-modal split between spatially coherent dynamics and spatially incoherent dynamics (Figure \ref{4}E). For the spatially coherent dynamics, the cumulative HPD was much lower than the spatially incoherent dynamics (median 24 cm, spatially coherent vs. median 238 cm, spatially incoherent).

@@

Our results indicate that most events with coherent spatial content are not dominated by the continuous dynamic, and thus correspond to trajectories that move relatively slowly compared to their purely continuous counteparts. We therefore examined these events in more detail. We first noted that most of the SWRs (16454 of 23382 classified SWRs or 70\%) were well described by a single dynamic. Surprisingly, the most common SWRs were categorized exclusively as stationary-continuous mixtures (9860 of 23382 classified SWRs or 42\%, Figure \ref{5}A). These events contain representations that move at rates more consistent with real-world movement (Fig. 1G), but were slightly shorter, on average, than events with continuous dynamics (median duration: stationary-continuous-mixture 72 ms vs. continuous 95 ms, Figure \ref{5}B). Nonetheless, both these slow-moving events and continuous events represented locations that were some distance away from the animal (mean trajectory distance from animal's position: stationary-continuous-mixture 51 cm vs. continuous 42 cm, Figure \ref{5}C). This indicates that the trajectories of these SWRs, like ones that have faster continuous dynamics, do not represent the animal's actual position and cannot be explained by erroneously detected SWRs.

@@

\section*{Discussion}
We developed a state space model that identifies and quantifies generative activity in terms of a mixture of three movement dynamics: stationary, continuous, and fragmented. This model is robust; it was relatively insensitive to misspecified transition parameters between dynamics and depended strongly on the hippocampal place representation structure. We show that this model is interpretable: it has well-defined parameters and allows us to explain a large fraction of SWRs in terms of these dynamics---far greater than had been explained before. This model is also flexible; it can be used with sorted or clusterless spikes, it can be used with 1D or 2D positions, and---because it uses a mixture of dynamics---it allows us to confidently discover not only SWRs with representations that move more or less linearly at a constant, high speed , but also representations that move non-linearly  with slower speeds or are stationary. These slower moving representations constitute the majority of representations seen during SWRs, but would not be found with the typical Bayesian decoder. The prevalence of these events indicates that hippocampal replay can recapitulate experience at both accelerated and real-world speeds.

Previous word reported that at most half of SWRs or high-population-activity events contained replay (REFS from above). This could lead one to the assumption that most events are not spatially coherent or meaningful. Our results indicate that this is not the case: in our dataset the large majority of SWRs (85\%) contained spatially coherent content, defined as content with some combination of stationary or continuous dynamics. These events included decoded representations that were highly spatial concentrated, confirming their spatial coherence. This was in stark contrast to unclassified events or events that included fragmented dynamics, which were associated with highly distributed decoded spatial representations.

Importantly, the spatially coherent events most often expressed a dynamic that was consistent with real-world movement speeds of <1 m/s. The next most common category were stationary events (Cite Yu etl al, Farooq et. al, perhaps Dupret et. al 2009?) that activated a representation of a single location, while the classical "replay" events made up only about ~19\% of classified events.

These results challenge the long-held notion that hippocampal replay trajectories necessarily proceed at many times the speed of the animal \cite{NadasdyReplayTimeCompression1999, LeeMemorySequentialExperience2002, DavidsonHippocampalReplayExtended2009}. Instead there is a much richer variety of speeds the animal can use to retrieve information about prior experience or generate possible future events, including slower moving representations that activate a spatially compact set of nearby locations and stationary representations of single locations. Interestingly, Farooq and Dragoi \cite{FarooqEmergencepreconfiguredplastic2019} reported that these stationary representations were most prevalent in juvenile rats and decreased to "chance" levels when the rats reached adulthood. Here, we show that they do not disappear, but are a very common occurrence with adult rats. Here we note that we chose to include a stationary distribution to explicitly evaluate the prevalence of stationary events in adult animals, but in principle a model with just continuous and fragmented dynamics could capture these events as well, as the continuous dynamic includes movement speeds of 0 m/s.

Our results may also explain the seeming conflict between Stella and colleagues \cite{StellaHippocampalReactivationRandom2019}, who reported that hippocampal replay trajectories followed Brownian-like diffusion dynamics, and Pfeiffer and Foster \cite{PfeifferAutoassociativedynamicsgeneration2015}, who reported that replay trajectories contained jumps in position. We found that while replay speeds are on average distributed log normally, there are still definitive "jumps" from one position to another in a small subset of the replays. Our model allows identification of both types of sequences. Our model also makes it possible to identify different dynamics within the same SWR. As the boundaries of each event are also arbitrary, we suggest that knowing the start and end of each dynamic may prove a powerful way to understand how inputs to the hippocampus influence these events and how these events in turn influence downstream structures.

Our identification of rich and varied dynamics in SWRs depended on several factors. First, we sought to describe as many SWRs as possible, instead of setting a high threshold for event detection. This more permissive approch was also critical for the identification of events that activate specific, immobility-associated locations in previous work (Cite Yu et. al). More broadly, there is no evidence that SWR and high population activity events constitute clearly distinct classes of events that can be identified with a single fixed threshold. Instead, the sizes of these events are well described by a unimodal, long-tailed distribution (cite cheng and frank 2008). High thresholds can therefore exclude large numbers of real events and potentially lead to erroneous conclusions.

Second, we used clusterless decoding, which enabled us to take advantage of information from neurons farther away from the tetrodes and not just those that could be cleanly and somewhat subjectively separated (REFS). Third, we built an explicit, but flexible model that allowed us to identify not just one dynamic, but multiple dynamics consistent with different speeds of motion. Fourth, our model avoids assumptions about temporal bins and does not make assumptions about the linear progression of events. Instead, the model can use any size of bin and allows for movement in any direction. Finally, we used an  acausal estimator of latent position, which allowed us to identify the specific timing of transitions in dynamics at a fine timescale.

Our model can be viewed as an extension to the multiple temporal models approach of Johnson and colleagues 2008 \cite{JohnsonMeasuringdistributedproperties2008}, which used model selection to determine the model with the most appropriate trajectory speed. Our model goes beyond the Johnson model in that it explicitly permits a mixture between the movement speeds, can work for arbitrary track geometries, and uses clusterless decoding. Our model also represents a middle ground between Hidden Markov-style models \cite{MaboudiUncoveringtemporalstructure2018, ChenBayesiannonparametricmethods2016, LindermanBayesiannonparametricapproach2016, ChenUncoveringspatialtopology2012}, which seek to be environment-agnostic discovers of sequential patterns and the typical Bayesian decoder, which typically use arbitrarily chosen bin sizes and  restrictive assumptions about the nature of the trajectories. In particular, our approach allows for a variety of dynamics while still yielding spatially interpretable results without the need to make assumptions about the right temporal bin size.

State-space models like the one we use here can enable a richer set of analyses of generative events that takes advantage of all of the available data. These approaches can be applied to not just SWRs but all times, providing a moment-by-moment estimate of the nature of the spatial representation in the hippocampus, including important information about the spatial coherence of that representation. The model can be extend to incorporate other previously experienced environments by training place field models on those environments and including the appropriate movement transition matrices for those environments. It can also be extended to account for trajectory task context and forward/reverse sequences as in \cite{DengRapidclassificationhippocampal2016}. Finally, models can be built to estimate any covariate, including movement direction (cite Kay 2020). We therefore suggest that this approach has the potential to provide a much richer understanding of neural representations throughout the brain.

@@

\subsection*{Simulated Data}
Encoding data for Figure \ref{1} and Figure \ref{fig:Figure1-Figure supplement 1} were generated by simulating 15 traversals of a 180 cm linear track. The spiking of 19 neurons was simulated using an inhomogeneous Poisson process where the instantaneous firing rate of each neuron changes according to a Gaussian place field. The Gaussian place fields had a variance of 36 cm and were spaced at 10 cm intervals over the 180 cm track. The decoding data for Figure \ref{1}G was generated by simulating 20,000 linear traversals of the 180 cm track, each with a unique constant speed, starting at 0.5 cm/s and increasing by 0.5 cm/s up to 10,000 cm/s. Each simulated neuron "spiked" when the traversal passed through the peak location of its place field.

@@

\subsection*{Behavioral Task}
All animals performed a W-track spatial alternation task, which is described in detail in \cite{KarlssonAwakereplayremote2009}. In brief, each day, animals alternated between 20 minute rest epochs in a rest box and 15 minutes run epochs in the W-track. On the W-track, animals were rewarded at the ends of an arm when that arm was the next correct arm in the sequence. Two rules determined the next correct arm. If the animal was in an outer arm, it had to run to the center arm. If the animal was in the center arm, it had to run from the center arm to the least recently visited outer arm. Animals were free to choose any arm at any time Only run epochs with at least 9 putative hippocampal pyramidal cells that fired at least 100 spikes were included in the analysis.

@@

\begin{equation*}
  p(x_{k} \mid x_{k-1}, I_{k}, I_{k-1}) = 
  \begin{blockarray}{*{3}{c} l}
    \begin{block}{*{3}{>{$\footnotesize}c<{$}} l}
      $I_{k} = Stationary$ & $I_{k} = Continuous$ & $I_{k} = Fragmented$ \\
    \end{block}
    \begin{block}{[*{3}{c}]>{$\footnotesize}l<{$}}
      \delta(x_{k-1}) & \mathcal{N}(x_{k-1}, 6.0) & \mathcal{U}(\min x, \max x) \bigstrut[t]& $I_{k-1} = Stationary$ \\
      \delta(x_{k-1}) & \mathcal{N}(x_{k-1}, 6.0)  & \mathcal{U}(\min x, \max x) & $I_{k-1} = Continuous$ \\
      \mathcal{U}(\min x, \max x) & \mathcal{U}(\min x, \max x) & \mathcal{U}(\min x, \max x) & $I_{k-1} = Fragmented$ \\
    \end{block}
  \end{blockarray}
\end{equation*}
where $\delta(x_{k-1})$ is an identity transition matrix where position cannot change from the previous time step, $\mathcal{N}(x_{k-1}, 6.0)$ is a random walk from the previous position with variance 6.0, and $\mathcal{U}(\min x, \max x)$ is a uniform transition that allows transitions to any possible position. As discussed in the Results, this means that when persisting in the same dynamic, the stationary, continuous, and fragmented dynamics are defined by the identity transition, the random walk, and the uniform transition, respectively. When transitioning to or from the fragmented dynamic, we assume we do not have any information about the position, so the transition is uniform. Finally, when the transition is from the stationary to continuous, we assume the position is spatially close where it was previously, so we use a random walk. When the transition is from continuous to stationary, we assume that the position is no longer changing, so we use the identity transition.

@@

\subsection*{Encoding - Clusterless}
In order to relate  each tetrode's unsorted spiking activity and waveform features  to position, we used kernel density estimation (KDE) to estimate the following distributions: $\pi(x)$, $p_{i}(x_k)$, and $p_{i}(x_k, \vec{m}_k)$. We used KDEs of the form: 
$$
kde(y) = \frac{1}{N h_1...h_D} \sum^{N}_{i=1} \prod^{D}_{d=1} K_{d}\left(\frac{y_d - y_{i,d}}{h_d}\right)
$$
where $y$ is the data with $D$ dimensions, $K$ is a 1 dimensional Gaussian kernel with bandwidth $h_d$, $y_{i, d}$ is the data observed during movement, $N$ is the number of observations during the movement period. For $\pi(x)$, $y_{i, d}$ is all positions observed during movement. For $p_{i}(x_k)$, $y_{i, d}$ is all positions at the time of multiunit spikes during movement. For $p_{i}(x_k, \vec{m}_k)$, $y_{i, d}$ is all positions at the time of multiunit spikes and their associated waveform features during movement. We choose the bandwidth $h_d$ to be 6.0 cm for any position dimension and 24.0 $\mu V$ for any spike amplitude dimension, because these parameters minimized decoding error in \cite{KloostermanBayesiandecodingusing2014}. The mean firing rate $\mu_k$ was also estimated during movement.
